%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\part{Proof Theory}\label{sec:proof_theory}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\emph{Proof Theory} is the study of Syntactic Consequence
(\S\ref{sec:syntactic_consequence}) by means of representing Deductive
Arguments (\S\ref{sec:logical_argument}) as Formal Proofs
(\S\ref{sec:formal_proof}).



% ====================================================================
\section{Formal System}\label{sec:formal_system}
% ====================================================================

A \emph{Formal System} (also \emph{Logical Calculus} or \emph{System
  of Inference}) is a Formal Language (\S\ref{sec:formal_language})
combined with a Deductive Apparatus (\S\ref{sec:deductive_apparatus})
in order to Derive Theorems (\S\ref{sec:formal_proof}).

In the context of a Formal System, Expressions that belong to the
underlying Formal Language are called ``\emph{Well-formed Formulas}''
(\emph{WFFs}) or just \emph{Formulas} (\S\ref{sec:formula}). The
actual Syntax will be determined within the bounds of the particular
Logical System (\S\ref{sec:logical_system}) under consideration.



% --------------------------------------------------------------------
\subsection{Deductive Apparatus} \label{sec:deductive_apparatus}
% --------------------------------------------------------------------

A \emph{Deductive Apparatus} (or \emph{Deductive System}) is a Set of
zero or more WFFs called \emph{Axioms} (\S\ref{sec:axiom}) and one or
more Functions on WFFs called \emph{Inference Rules}
(\S\ref{sec:inference_rule}) that can be used to Derive Theorems
(\S\ref{sec:formal_proof}).

Deduction (\S\ref{sec:deductive_inference}) proper is the top-down,
Reductive (\S\ref{sec:abstract_rewrite}) process that starts with
general Axioms and Reduces to the specific Theorem that is being
proved by repeated application of Inference Rules.

A Deductive System gives a precise account of Syntactic Consequence
(\S\ref{sec:syntactic_consequence}) with respect to a Formal Language.
A Formal System is termed \emph{Effective} (i.e. Recursively
Definable) if the Set of Axioms and the Set of Inference Rules are
Decidable or Semi-decidable (\S\ref{sec:decidable_set}). The notion of
``Theorem'', however, is not in general Effective.



% --------------------------------------------------------------------
\subsection{Inference Rule} \label{sec:inference_rule}
% --------------------------------------------------------------------

An \emph{Inference Rule} (or \emph{Transformation Rule}) is a Function
which takes an input Set of WFFs called \emph{Premises} and returns an
output Set of WFFs called the \emph{Conclusion}. An alternative
formulation for Inference Rules is as a \emph{Deducibility Relation},
$\vdash$, that holds between zero or more Premises and one or more
Conclusions.

Taken individually, both Premises and Conclusions are Propositions
(\S\ref{sec:proposition}), that is, bearers of Truth Value
(\S\ref{sec:truth_value}). The Conclusion relies on the truth of the
Premises; if the Premises are left Unsatisfied
(\S\ref{sec:satisfaction}), then the Derivation is \emph{Hypothetical}
and the Premises \emph{Hypotheses}.

Inference Rules may be identified as redundant in two senses:
\begin{enumerate}
\item An \emph{Admissible Rule} is one which does not change the Set
  of Theorems in a Formal System when it is added.
\item A \emph{Derivable Rule} is a case of an Admissible Rule that has
  been Derived from existing rules.
\end{enumerate}



\subsubsection{Rule Schema}\label{sec:rule_schema}



% --------------------------------------------------------------------
\subsection{Axiom}\label{sec:axiom}
% --------------------------------------------------------------------

\emph{Axioms} are special cases of Inference Rules which have no
Premises, only a universally Valid (\S\ref{sec:validity}) Conclusion;
that is, a \emph{Logical Assertion} (\S\ref{sec:sequent_notation}).
Axioms may be further differentiated from Rules by saying that Rules
are statements \emph{about} the System, and Axioms are statements
\emph{in} the System.

Axioms may be divided into two kinds: \emph{Logical Axioms} of a
tautological sort, and \emph{Non-logical Axioms}, hereafter referred
to as \emph{Postulates}, that play the role of assumptions: defining
properties of the domain of the Theory (\S\ref{sec:formal_theory}) in
question.

Systems with a large number of Axioms and few other Inference Rules
are called \emph{Axiomatic Systems} (\S\ref{sec:axiomatic_system}).



\subsubsection{Logical Axiom}\label{sec:logical_axiom}



\subsubsection{Non-logical Axiom}\label{sec:nonlogical_axiom}

\emph{Non-logical Axiom} (or \emph{Postulate})



\subsubsection{Axiom Schema}\label{sec:axiom_schema}

An \emph{Axiom Schema} is a template for Axioms in which one or more
Schematic Variables (\S\ref{sec:metalanguage}) appear, standing for a
Subformula in the Object Language of the System. For a Language with
infinitely many WFFs, an Axiom Schema describes a countably infinite
number of Axioms. A System without Schema is termed \emph{Finitely
  Axiomatized}.



% ====================================================================
\section{Formal Proof} \label{sec:formal_proof}
% ====================================================================

A \emph{Formal Proof} (or \emph{Derivation}) is a finite sequence of
Well-formed Formulas of a Formal System that are either Axioms or
follow from the preceding Formulas by an Inference Rule. A Formal
Proof is said to provide \emph{Justification} for the Theorem
(\S\ref{sec:theorem}) thus Derived.

For a Set of WFFs, $\Gamma$, in a Formal System, $\mathcal{S}$, there
is a Syntactic Consequence, $A$, if there is a Formal Proof of $A$
from the Set $\Gamma$:
\[
    \Gamma \vdash_{\mathcal{S}} A
\]



% --------------------------------------------------------------------
\subsection{Theorem} \label{sec:theorem}
% --------------------------------------------------------------------

The concluding Formula in the sequence is a \emph{Theorem}. A Theorem
is a Syntactic Consequence (\S\ref{sec:syntactic_consequence}) of the
preceding WFFs in the Proof. A Theorem used in the course of Deriving
a further Theorem is called a \emph{Lemma}.



% --------------------------------------------------------------------
\subsection{Analytic Proof} \label{sec:analytic_proof}
% --------------------------------------------------------------------

A \emph{Theoretic Analytic Proof} begins with an assumption and
proceeds to an accepted truth (an Axiom, or contradiction as in
\S\ref{sec:tableau_calculus} Analytic Tableau).



% --------------------------------------------------------------------
\subsection{Synthetic Proof} \label{sec:synthetic_proof}
% --------------------------------------------------------------------

A \emph{Synthetic Proof} is the reverse of this process; beginning
with known truths and reasoning up to the desired Proof. A
\emph{Problematic Analytic Proof} is constructed from given conditions
that are to be satisfied.



% --------------------------------------------------------------------
\subsection{Constructive Proof}\label{sec:constructive_proof}
% --------------------------------------------------------------------

\emph{Existence Theorem}



\subsubsection{Realizability} \label{sec:realizability}



% --------------------------------------------------------------------
\subsection{Non-constructive Proof}\label{sec:nonconstructive_proof}
% --------------------------------------------------------------------



% --------------------------------------------------------------------
\subsection{Proof Procedure}\label{sec:proof_procedure}
% --------------------------------------------------------------------

\subsubsection{Direct Proof}\label{sec:direct_proof}

\subsubsection{Indirect Proof}\label{sec:indirect_proof}

\subsubsection{Reverse Mathematics}\label{sec:reverse_mathematics}



% ====================================================================
\section{Formal Theory}\label{sec:formal_theory}
% ====================================================================

For a Formal Language (\S\ref{sec:formal_language}), $L$, a Formal
Theory, $\mathcal{T}$, is a Subset of a Conceptual Class of
\emph{Elementary Expressions} (\emph{Elementary Statements} or
\emph{Well-formed Formulas} \S\ref{sec:formula}), $\mathcal{E}$,
consisting of \emph{Elementary Theorems} considered to be True. In a
\emph{Deductive Theory} (\S\ref{sec:deductive_theory}), $\mathcal{T}$
is an \emph{Inductive Class} such that some of the Elementary Theorems
are taken to be \emph{Axioms} and any Expressions Deducible by those
Axioms are also Elementary Theorems in $\mathcal{T}$.



% --------------------------------------------------------------------
\subsection{Deductive Theory}\label{sec:deductive_theory}
% --------------------------------------------------------------------

In the context of a Deductive System, $\mathcal{S}$, a \emph{Deductive
  Theory} (or \emph{Inductive Class}), $T$, is a Formal Theory that
consists of all the Theorems that are Derivable in $\mathcal{S}$,
including any Axioms.



% --------------------------------------------------------------------
\subsection{Subtheory \& Extension}\label{sec:subtheory}
% --------------------------------------------------------------------

A Theory $S$ is a \emph{Subtheory} of a Theory $T$ if $S \subseteq T$.
$T$ is then a \emph{Supertheory} (or \emph{Extension}) of $S$.



% --------------------------------------------------------------------
\subsection{Conservative Extension}\label{sec:conservative_extension}
% --------------------------------------------------------------------

Given two Theories $T_1$ and $T_2$ with underlying Formal Languages
$L_1$ and $L_2$ respectively, $T_2$ is a \emph{Conservative Extension}
of $T_1$ if:
\begin{enumerate}
    \item $L_2$ extends $L_1$
    \item $T_1 \subseteq T_2$
    \item any Theorem of $T_2$ in $L_1$ is already a Theorem of $T_1$
\end{enumerate}
An Extension which is not Conservative may be called a \emph{Proper
  Extension}.

See also \emph{Conservativity} (\S\ref{sec:conservativity})



% --------------------------------------------------------------------
\subsection{Consistent Theory}\label{sec:consistent_theory}
% --------------------------------------------------------------------

A Theory is \emph{Absolutely Consistent} (also \emph{Coherent} or
\emph{Non-trivial}) if not every WFF in the underlying Language is a
Theorem.

A Non-trivial Theory is \emph{Consistent} if it satisfies the
\emph{Principle of Explosion}, that is, Proof of a contradiction
Implies Proof of all other WFFs. Therefore, a Consistent Theory must
be free of contradictions.

The Conservative Extension of a Consistent Theory is Consistent.

See also \emph{Paraconsistency} (\S\ref{sec:paraconsistent_inference}).



\subsubsection{$\omega$-consistent}\label{sec:omega_consistent}



% --------------------------------------------------------------------
\subsection{Complete}\label{sec:complete_theory}
% --------------------------------------------------------------------

A Consistent Theory is said to be \emph{Complete} if for every
Sentence in the Language of that Theory, the Theory contains either
that Sentence or its Negation. Such a Set of Sentences is a
\emph{Maximal Consistent Set}.



% --------------------------------------------------------------------
\subsection{Satisfiable}\label{sec:satisfiable_theory}
% --------------------------------------------------------------------

A Theory is \emph{Satisfiable} if there exists a Model
(\S\ref{sec:structure}) for that Theory.

A Theory is \emph{Categorical} if it has just one Model up to
Isomorphism.

%FIXME connections with stability theory

By G\"odels Completeness Theorem, a First-order
(\S\ref{sec:predicate_logic}) Theory is Consistent if and only if it
is Satisfiable. Higher-order Logics may allow Syntactically Consistent
Theories that are not Satisfiable.



% --------------------------------------------------------------------
\subsection{Sound}\label{sec:sound_theory}
% --------------------------------------------------------------------

A Theory is \emph{Sound} when every Theorem is Valid
(\S\ref{sec:validity}).



% --------------------------------------------------------------------
\subsection{Effective}\label{sec:effective_theory}
% --------------------------------------------------------------------

A Theory is \emph{Effective} when it is possible to Effectively
(\S\ref{sec:effective_method}) determine whether a Deduction is Valid
or not.



% --------------------------------------------------------------------
\subsection{Synthetic Theory}\label{sec:synthetic_theory}
% --------------------------------------------------------------------

% --------------------------------------------------------------------
\subsection{Duality Principle}\label{sec:duality_principle}
% --------------------------------------------------------------------



% ====================================================================
\section{Proof Calculus}\label{sec:proof_calculus}
% ====================================================================

A \emph{Proof Calculus} is a family of Formal Systems
(\S\ref{sec:formal_system}) that use a common style of Formal
Inference for Inference Rules.



% --------------------------------------------------------------------
\subsection{Axiomatic System} \label{sec:axiomatic_system}
% --------------------------------------------------------------------

\emph{Axiomatic System}

In contrast to Sequent Calculus (\S\ref{sec:sequent_calculus}) or
Natural Deduction (\S\ref{sec:natural_deduction}), Axiomatic Systems
are more useful in the context of Model Theory (Part
\ref{sec:model_theory}).



\subsubsection{Hilbert Systems}\label{sec:hilbert_systems}

\emph{Hilbert Systems} are characterized by having a large number of
Axiom Schema and few Inference Rules-- just Modus Ponens for
Propositional Logics and Universal Generalization for Predicate
Logic. In a Hilbert System, Judgements and Formulas are not
differentiated. A Theorem in a Hilbert System is the Concluding
Judgement in a Derivation.

A Hilbert System is differentiated from Systems of Natural Deduction
by not having any Rules that change the Context of a Formula.



% --------------------------------------------------------------------
\subsection{Structural Proof Theory}\label{sec:structural_proof}
% --------------------------------------------------------------------

\emph{Structural Proof Theory} studies Proof Calculi that support
Analytic Proof (\S\ref{sec:analytic_proof}); that is Proofs that are
\emph{Cut-free} (they do not use the \emph{Cut Rule}) or in Normal
Form (\S\ref{sec:normal_form}).



\subsubsection{Natural Deduction}\label{sec:natural_deduction}
\cite{prawitz65}

\emph{Natural Deduction} provides a generalization of Formal Proofs.
Contrasted with Axiomatic Systems (\S\ref{sec:axiomatic_system}),
Systems of Natural Deduction include many Inference Rules but few or
no Axioms. A Natural Deduction System allows Judgements with multiple
Antecedents and a single Succedent:
\[
    A_1,\ldots,A_n \vdash B
\]
Inference Rules in Natural Deduction have the general notation
\[
    {
        \frac{J_1 \quad J_2 \quad \cdots \quad J_n}
        {J}
    } name
\]
where the Rule with name $name$ has Premises of zero or more
Judgements $J_i$ and the Judgement $J$ is the Conclusion.

Inference Rules that introduce a Logical Connective in the Conclusion
are called \emph{Introduction Rules}. Example
\[
    {
        \frac{A\;\mathrm{true} \quad B\;\mathrm{true}}
        {(A \wedge B)\;\mathrm{true}}
    } \wedge_I
\]
where $A$ and $B$ are Propositions.

Conversely, Inference Rules that remove Logical Connectives are
\emph{Elimination Rules}.
\[
    {
        \frac{A \wedge B\;\mathrm{true}}
        {A\;\mathrm{true}}
    } \wedge_E
\]

\emph{Hypothetical Derivations} (reasoning from \emph{Assumptions})
are required for Implication Introduction or Disjunction
Elimination. The general form of a Hypothetical Derivation with
Antecedents $D_i$ and Succedent $J$:
\[
    D_1 \quad D_2 \cdots D_n
\]\[
    \vdots
\]\[
    J
\]
Introduction Rules for Implication:
\[
    {
        \frac{}
        {A\;\mathrm{true}}
    } u
\]\[
    \vdots
\]\[
    {
        \frac{B\;\mathrm{true}}
        {A \rightarrow B\;\mathrm{true}}
    } \rightarrow_{I^u}
\]
The Premise $u$ here is considered \emph{discharged} by the Rule
$I^u$; that is the scope of $u$ does not extend past $I^u$.
Elimination Rule for Implication (Modus Ponens):
\[
    {
        \frac{A \rightarrow B\;\mathrm{true} \quad A\;\mathrm{true}}
        {B\;\mathrm{true}}
    } \rightarrow_{E}
\]
Disjunctive Elimination:
\[
    \frac{
    A \vee B\;\mathrm{true} \quad
    \begin{matrix}
        {
            \frac{}
            {A\;\mathrm{true}}
        }u \\
        \vdots \\
        C\;\mathrm{true}
    \end{matrix}
    \quad
    \begin{matrix}
        {
            \frac{}
            {B\;\mathrm{true}}
        }w \\
        \vdots \\
        C\;\mathrm{true}
    \end{matrix}
    }{ C\;\mathrm{true}}\wedge_{E^{u,w}}
\]
A Theory is \emph{Locally Consistent} (or \emph{Locally Reducible}) if
an Introduction of a Connective followed by its Elimination can be
equivalently Derived without these steps. The dual to Local
Consistency is \emph{Local Completeness} which states that Elimination
rules can decompose a Connective into the forms of its Introduction
Rule. These correspond to $\beta$-reduction and $\eta$-conversion in
$\lambda$-Calculus (\S\ref{sec:lambda_calculus}) where Propositions
are \emph{Types} and Proofs are \emph{Programs} (see
\emph{Curry-Howard Correspondence} \S\ref{sec:curry_howard}).

If an entire Derivation has only Eliminations followed by
Introductions, it is said to be in \emph{Normal Form}
(\S\ref{sec:normal_form}).

In a Formal Proof, the Judgements representing Antecedents are
presented as Rules with no Premises, named by a \emph{Proof Variable}
(from a countable set $V$ of variables):
\[
    \frac{}{J_1}u_1 \; \frac{}{J_2}u_2 \; \cdots \frac{}{J_n}u_n
\]\[
    \vdots
\]\[
    J
\]
where $u_i \in V$. Written in Sequent Notation:
\[
    u_1:J_1, u_2:J_2, \ldots, u_n:J_n \vdash J
\]
This convention is sometimes called \emph{Localized Hypotheses}. In
general, $\pi : A$ may be read ``$\pi$ is a proof of $A$''.



\subsubsection{Sequent Calculus}\label{sec:sequent_calculus}

In \emph{Sequent Calculus} (or \emph{Gentzen System}) a Formal Proof
is a Sequence of Sequents (\S\ref{sec:sequent_notation}) where each
successive Sequent is Derivable from prior Sequents by Inference
Rules.



\paragraph{Sequent Notation}\label{sec:sequent_notation} \hfill \\

A \emph{Sequent} is a specific kind of Judgement
(\S\ref{sec:judgement}) of the form
\[
    \Gamma \vdash \Sigma
\]
where the \emph{Antecedent}, $\Gamma$, is a Conjunctive sequence of
Formulas, and the \emph{Succedent}, $\Sigma$, is a Disjunctive
sequence of Formulas. Together, Antecedents and Succedents are
\emph{Cedents}. The Extra-logical Operators, $\vdash$, and $,$
(comma), are called \emph{Structural Operators} and Rules which change
only Structural Operators are \emph{Structural Rules} (as opposed to
\emph{Logical Rules}). A sequence of Cedents may be called a
\emph{Context}, but \emph{the} Context for a specific Judgement is
usually meant to be the Antecedent.

\emph{Weakening} refers to a Rule that introduces arbitrary elements
to a Sequent. \emph{Contraction} refers to a Rule that removes
multiple occurences of some element and \emph{Permutation} refers to
the re-ordering of elements. Logics lacking Structural Rules are
called \emph{Substructural Logics} (\S\ref{sec:substructural_logic}).
If Sequents are defined as Sets or Multisets (\S\ref{sec:multiset})
instead of Sequences (that is, unordered Sets), then the Permutation
rule is obsolete, likewise the Contraction Rule would be obsolete for
Sets instead of Sequences.

In a general Sequent Calculus there may be any number of Formulas on
either side
\[
    A_1, \ldots, A_n \vdash B_1, \ldots, B_k
\]
is equivalent to
\[
    \vdash(A_1 \wedge \cdots \wedge A_n) \rightarrow (B_1 \vee \cdots \vee B_k)
\]
and the dual nature of Judgements and negation can be expressed by the
dual forms
\[
    \vdash \neg A_1 \vee \cdots \vee \neg A_n \vee B_1 \vee \cdots
    \vee B_k
\]
and
\[
    \vdash \neg(A_1 \wedge \cdots \wedge A_n \wedge \neg B_1 \wedge
    \cdots \wedge \neg B_k)
\]

A Sequent with no Succedent ($\Gamma \vdash$) is a
\emph{Contradiction} meaning it proves falsity which is
Inconsistent. A Sequent with no Antecedent ($\vdash \Sigma$) is a
\emph{Logical Assertion} and the Succedent is a
\emph{Tautology}. Theorems are those of the form $\vdash B$ which are
the Conclusion of a Valid Proof.



\paragraph{Cut Elimination}\label{sec:cut_elimination} \hfill \\

The Rule for \emph{Cut} is as follows:
\[
    \frac{
        \Gamma \vdash \Delta, A \quad A, \Sigma \vdash \Pi
    }{
        \Gamma, \Sigma \vdash \Delta, \Pi
    }(Cut)
\]
It states that when a Formula $A$ that can be Concluded can also be
used as a Premise, it can be \emph{cut} out and the Derivations joined
together. That is, wherever the Lemma $A$ occurs, it can be
substituted for the Proof of $A$. This means that the Cut Rule is an
Admissible Rule (\S\ref{sec:inference_rule}).

The \emph{Cut-elimination Theorem} states that any Judgement with a
Proof in Sequent Calculus that uses the Cut Rule may be expressed as a
\emph{Cut-free} Proof without using the Cut Rule. Usually,
demonstrating the existence of the Cut-elmination Theorem implies that
the System is Consistent since that would rule-out the possibility of
Proof of Contradiction.



\paragraph{$\mathbf{LK}$}\label{sec:lk} \hfill \\

Formalization of Classical Logic (\S\ref{sec:classical_logic})
(sound and complete in First-Order) with Sequents having zero or more
RHS Formulas. Allowing multiple RHS Formulas with a \emph{Right
  Contraction Rule} is equivalent to the admissibility of the
\emph{Law of the Excluded Middle}.



\paragraph{$\mathbf{LJ}$}\label{sec:lj} \hfill \\

Formalization of Intuitionistic Logic
(\S\ref{sec:intuitionistic_logic}) obtained from modifying
$\mathbf{LK}$ to allow at most one RHS Formula in a Sequent. The Cut
Rule for $\mathbf{LJ}$:
\[
    \frac{
        \Gamma \vdash A \quad \Pi, A \vdash B
    }{
        \Gamma, \Pi \vdash B
    }(Cut)
\]



\paragraph{Substructural Rule Sets}\label{sec:substructural_rule}
\hfill \\

A Substructural Logic (\S\ref{sec:substructural_logic}) lacking the
usual Structural Rules is usually weaker than $\mathbf{LK}$.

In Relevance Logic (\S\ref{sec:relevance_logic}), Weakening Rules
are not included on the grounds that introduced Formulas are not
\emph{Relevant}.

In Linear Logic (\S\ref{sec:linear_logic}), duplicate Formulas are
treated differently so Contraction and Weakening Rules are is absent
or controlled.



\subsubsection{Calculus of Structures}\label{sec:calculus_of_structures}

\emph{Calculus of Structures} is a Proof Calculus with \emph{Deep
  Inference} for studying Non-commutative Logic
(\S\ref{sec:noncommutative_logic}). Deep Inference is a
generalization of Structure to handle greater Structural complexity
\cite{schutte77}.



% --------------------------------------------------------------------
\subsection{Tableau Calculus}\label{sec:tableau_calculus}
% --------------------------------------------------------------------

\emph{Tableau Calculus} (or \emph{Method of Analytic Tableau}) is
commonly used as a Proof procedure for Modal Logics
(\S\ref{sec:modal_logic}). An Analytic Tableau is a tree with a
Formula at the root and a Subformula at each node. A specific Tableau
Calculus is a finite collection of Rules for breaking down Logical
Connectives into constituent parts. Rules can be expressed as Sets,
Multisets, Lists, or Trees of Formulas. If Sets of Formulas are used
at each node (\emph{Set-labeled Tableau}), they are taken in
Conjunction.

A \emph{Strongly Complete Tableau} is one in which every Formula in
every branch has been expanded.



\subsubsection{Refutation Tableau}\label{sec:refutation_tableau}

A \emph{Refutation Tableau} attempts to show that a negation of the
root Formula cannot be satisfied, thereby proving Logical Truth of the
Formula. Rules for handling Logical Connectives may produce a branch
in the tree and if a branch leads to a Contradiction, the branch is
closed and if all branches are closed the Proof is complete and the
root Formula is proved. Nodes on a single branch are considered in
Conjunction, Nodes on separate branches are considered Disjunctively.



\subsubsection{Destructive \& Non-destructive Tableau}
\label{sec:destructive_tableau}

\emph{Non-destructive Tableau Calculi} use Rules that only allow
addition of nodes, while \emph{Destructive Tableau Calculi} use Rules
that allow modification of existing nodes.



\subsubsection{Proof Confluence} \label{sec:proof_confluence}

\emph{Proof Confluence} is the property of a Tableau Calculus that a
closed Tableau (for an un-satisfiable set of Propositions) can always
be generated from an arbitrary partially constructed Tableau
regardless of which Rules are chosen at each application (if a choice
between Rules is available).



\subsubsection{Unification} \label{sec:unification}

A method of dealing with non-determinism of Rules involving Universal
Quantification (in First-order Tableau) is called \emph{Unification}.
This allows Free Variables to be substituted in the Rule for
Eliminating Universal Quantifiers, which can later be Unified by
choosing an appropriate Term to close the branch.



\subsubsection{Clause Tabeau} \label{sec:clause_tableau}

\emph{Clause Tableau} (Tableau Method applied to sets of Clauses) may
be used for increased efficiency.



\paragraph{Connection Tableau} \label{sec:connection_tableau}
\hfill \\

\emph{Connection Tableau} restrict expansion of Clause Tableau
branches (not the bare root) to contain only Literals that unify with
a Literal already on the branch (\emph{Weak Connectedness}) or a
Literal in the current leaf (\emph{Strong Connectedness}).



% --------------------------------------------------------------------
\subsection{Syllogistic Calculus} \label{sec:syllogistic_calculus}
% --------------------------------------------------------------------



% ====================================================================
\section{Logical System}\label{sec:logical_system}
% ====================================================================

A \emph{Logical System} (or \emph{System of Logic}) is a Formal System
(\S\ref{sec:formal_system}) together with a form of Semantics that
gives Meanings to the Formulas in the underlying Language.

A Structure (\S\ref{sec:structure}) that Satisfies
(\S\ref{sec:satisfaction}) the Axioms of a Logical System is called a
\emph{Model} (\S\ref{sec:model}) of the System.



% --------------------------------------------------------------------
\subsection{Classical Logic}\label{sec:classical_logic}
% --------------------------------------------------------------------

\emph{Classical Logic} is the class of Propositional and
First-order Systems of Logic characterized by the following Inference
Rules:

\begin{description}

\item [Tertium non datur] (\emph{Law of excluded middle})
    \[\vdash(p \vee \neg p)\]

\item [Double Negation]
    \[p \vdash \neg\neg p\]

\item [Law of Non-contradiction]
    \[\vdash \neg(p \wedge \neg p)\]

\item [Ex falso quodlibet] (\emph{Principle of explosion},
  \emph{Principle of Psuedo-Scotus})
    \[\vdash 0 \rightarrow p\]

%FIXME finish properties and rules

\end{description}

The intended Semantic Interpretation (\S\ref{sec:interpretation}) of
Classical Logic is subject to the Principle of Bivalence
(\S\ref{sec:bivalence_principal}) which says that every Proposition
has one Truth-value: True or False. Non-classical Logics such as
Intuitionistic Logic (\S\ref{sec:intuitionistic_logic}) does not have
this Property.



% --------------------------------------------------------------------
\subsection{Intuitionistic Logic}\label{sec:intuitionistic_logic}
% --------------------------------------------------------------------

\emph{Intuitionistic Logic} (or \emph{Constructive Logic}) replaces
Truth with the concept of \emph{Constructive Provability}. This is to
say that Logical Operations in Intuitionistic Logic preserve
\emph{Justification} (\S\ref{sec:formal_proof}) rather than
Truth-value. Such systems are restrictions of Classical Logic
(\S\ref{sec:classical_logic}) without the Law of the Excluded Middle
or Double Negation Elimination.

Whereas First-order Logic (\S\ref{sec:firstorder_logic}) is a
foundation for Set Theory (Part \ref{sec:set_theory}), Intuitionistic
Logic is used as a foundation for Type Theory (Part
\ref{sec:type_theory}) and Constructive Set Theory
(\S\ref{sec:constructive_set_theory}).

A Formula in Intuitionistic Logic does not necessarily have a Prenex
Normal Form (\S\ref{sec:prenex_normal}).

Intuitionistic Logic is Modelled (\S\ref{sec:model}) by Heyting
Algebra (\S\ref{sec:heyting_semantics}) or Kripke Semantics
(\S\ref{sec:kripke_semantics}) and lacks the Principle of Bivalence
(\S\ref{sec:bivalence_principal}); thus there is no Sole-sufficient
Operator in Intuitionistic Logic. A Formula in Intuitionistic Logic is
Valid if and only if it receives the Value of the Top Element for any
Valuation on a Heyting Algebra.

Functionally Complete Sets of Operators are:
\[
    \{ \vee, \leftrightarrow, \bot \}
\]
and:
\[
    \{ \vee, \leftrightarrow, \neg \}
\]



\subsubsection{Existence Property}\label{sec:existence_property}

Proofs in a Theory (\S\ref{sec:formal_theory}), $\mathcal{T}$, based
on Intuitionistic Logic have the \emph{Existence Property}:
\[
    (\exists x)A(x) \in \mathcal{T} \rightarrow (\exists t)A(t)
\]
where $A(x)$ has $x$ as the only Free Variable and $t$ is a Term.



\subsubsection{Disjunction Property}\label{sec:disjunction_property}

\emph{Disjunction Property}:
\[
    A \vee B \in \mathcal{T} \rightarrow A \in \mathcal{T} \vee B \in \mathcal{T}
\]



\subsubsection{Minimal Logic}\label{sec:minimal_logic}

\emph{Minimal Logic} is an Intuitionistic Logic without the Principle
of Explosion (\S\ref{sec:explosion_principle}).



\subsubsection{Combinatory Logic}\label{sec:combinatory_logic}

\emph{Combinator}



\subsubsection{Intermediate Logic}\label{sec:intermediate_logic}

\emph{Intermediate Logic} is an extended Intuitionistic Logic
(\emph{Superintuitionistic Logic}) that is Consistent and still weaker
than the strongest Consistent Superintuitionistic Logic: Classical
Logic.



% --------------------------------------------------------------------
\subsection{Substructural Logic}\label{sec:substructural_logic}
% --------------------------------------------------------------------

\subsubsection{Relevance Logic}\label{sec:relevance_logic}

\emph{Relevance Logic} (or \emph{Relevant Logic})



\subsubsection{Linear Logic}\label{sec:linear_logic}

\paragraph{Noncommutative Logic}\label{sec:noncommutative_logic}
\hfill \\

\emph{Ordered Logic}



% --------------------------------------------------------------------
\subsection{Ordinal Logic}\label{sec:ordinal_logic}
% --------------------------------------------------------------------

Alan Turing's PhD Thesis \cite{turing38}



% --------------------------------------------------------------------
\subsection{Categorical Logic}\label{sec:categorical_logic}
% --------------------------------------------------------------------

% --------------------------------------------------------------------
\subsection{Hoare Logic}\label{sec:hoare_logic}
% --------------------------------------------------------------------

\emph{Concurrency}



% --------------------------------------------------------------------
\subsection{Free Logic}\label{sec:free_logic}
% --------------------------------------------------------------------

\subsubsection{Inclusive Logic}\label{sec:inclusive_logic}



% --------------------------------------------------------------------
\subsection{Connexive Logic}\label{sec:connexive_logic}
% --------------------------------------------------------------------

% --------------------------------------------------------------------
\subsection{Non-monotonic Logic}\label{sec:nonmonotonic_logic}
% --------------------------------------------------------------------

Defeasible Inference (\S\ref{sec:defeasible_inference})



\subsubsection{Default Logic}\label{sec:default_logic}

Default Inference (\S\ref{sec:default_inference})



\subsubsection{Autoepistemic Logic}\label{sec:autoepistemic_logic}



% --------------------------------------------------------------------
\subsection{Paraconsistent Logic}\label{sec:paraconsistent_logic}
% --------------------------------------------------------------------

\subsubsection{Discursive Logic}\label{sec:discursive_logic}

\subsubsection{Non-adjunctive Logic}\label{sec:nonadjunctive_logic}

\subsubsection{Adaptive Logic}\label{sec:adaptive_logic}

\subsubsection{Formal Inconsistency}\label{sec:formal_inconsistency}

\subsubsection{Logic of Paradox}\label{sec:logic_of_paradox}

Graham Priest



% ====================================================================
\section{Metatheory} \label{sec:metatheory}
% ====================================================================

A \emph{Metatheory} is a Theory whose subject matter is some other
Theory.



% --------------------------------------------------------------------
\subsection{Syntactic Consequence}\label{sec:syntactic_consequence}
% --------------------------------------------------------------------

For a Formal System, $\mathcal{S}$, a WFF, $A$, is a \emph{Syntactic
  Consequence} of a possibly empty Set of Formulas, $\Gamma$, if there
is a Formal Proof (\S\ref{sec:formal_proof}) of $A$ from $\Gamma$ in
$\mathcal{S}$:
\[
    \Gamma \vdash_{\mathcal{S}} A
\]
where $\vdash_{\mathcal{S}}$ is the \emph{Syntactic Consequence
  Relation} on $\mathcal{S}$.



% --------------------------------------------------------------------
\subsection{Metatheorem}\label{sec:metatheorem}
% --------------------------------------------------------------------

A \emph{Metatheorem} is a Theorem proved within a Metatheory.



% --------------------------------------------------------------------
\subsection{Judgement}\label{sec:judgement}
% --------------------------------------------------------------------

A \emph{Judgement} is an inductively definable assertion in the
Metatheory of a Logical System. That is, one that includes
Extra-logical Symbols, namely that of Logical Consequence, $\vdash$,
and commas used in Sequents (\S\ref{sec:sequent_notation}). In this
way, Axioms are Judgements, and a Formal Proof expresses a Judgement
with the Premises being a sequence of Judgements and the Conclusion
also a Judgement.
